---
title: "R Coding Assignment 1"
author: "Aiden Jajo"
subtitle: "San Diego State University - STAT550"
output:
  pdf_document: default
  word_document: default
---

### R A First Look at Simulation

This section of the text introduces the concept of Monte Carlo simulation, along with some R commands using two examples (which are provided below or as script files). The first code starts with Example 1.32.
```{r example 1}
n <- 100 #set for example

# Generate random binary outcomes (0 or 1) simulating coin flips
sample(0:1, n, replace = TRUE)

# Calculate mean - theoretical expectation is 0.5 for fair coin
mean(sample(0:1, n, replace = TRUE))
```

### try creating a sample now

use the code above and make a sample of sampling from the digits 0-9 and find the mean.
```{r Code here 1}
n <- 100

# Sample from digits 0-9 with replacement
sample_0_to_9 <- sample(0:9, n, replace = TRUE)
sample_0_to_9

# Calculate mean - expected value is 4.5
mean(sample_0_to_9)
```

### Here are some of the functions that will be useful as we go through the course.

choose(n,k) - computes combination of choosing k objects from n. 
prod(n:m) - computes the product of values from n to m. To do a permutation of r objects out of n, you want to use prod(n:n-r+1). (For whatever reason, this doesn't work if you assign n and r first, just put in the numbers). 
factorial(x) - computes x!
```{r example 2}
# Combinations: order doesn't matter
choose(n = 10, k = 2)
## [1] 45

# Permutations: order matters
prod(8:6) # permutation of 3 objects out of 8
## [1] 336

# Factorial: total arrangements
factorial(x = 4)
##[1] 24
```

## Try the following

Find how many ways you can select 10 things from a group of 30

Find the total possible ways to arrange 10 people in a line for a photo.

Find 9!
```{r practice}
# Combination: selecting 10 from 30
choose(30, 10)

# Permutation: arranging 10 people in line
factorial(10)

# Factorial calculation
factorial(9)
```