---
title: "R Coding Assignment 2: BMW Viewing Interest Simulation"
author: "Aiden Jajo"
subtitle: "San Diego State University - STAT550"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

The BMW of Carlsbad dealership wants to know which BMW models have the most interest from customer requests to view each vehicle. The models available at the BMW dealership are:

- 2 Series
- 3 Series
- 4 Series
- 5 Series
- 7 Series
- i4
- i5
- i7
- ix
- x2
- x3
- x5
- x6
- x7
- XM

In order to estimate the interest in the BMW models from customers, the simulation will sample 1000 customer requests to view the models available at the BMW Carlsbad location.
```{r BMW Models}
# Define vector of all BMW models available at the dealership
bmw_models = c("2 Series", 
               "3 Series", 
               "4 Series", 
               "5 Series", 
               "7 Series", 
               "i4", 
               "i5", 
               "i7", 
               "ix", 
               "x2", 
               "x3", 
               "x5", 
               "x6", 
               "x7", 
               "XM")

# Initialize empty vector to store simulated customer requests
view_intrest = c()

# Simulate 1000 customer viewing requests
# Each iteration randomly selects one BMW model with equal probability
for (i in 1:1000) {
  model = sample(bmw_models, 1)
  view_intrest = c(view_intrest, model)
}

# Create frequency table showing number of requests per model
# Uses kable for formatted table output
library(knitr)
kable(as.data.frame(table(view_intrest)), col.names = c("Model", "Requests"))
```